TKT RNAseq Experiment: Differential Expression Analysis

### 2023-06-21 TKT_RNAseq/R Code/EdgeR_Analysis.Rmd
TKT edgeR analysis update

### Male and female samples processed together with TEs and outlier libraries removed
- #### Figure \@ref(fig:fig3): MDS plot, all samples

- #### Figure \@ref(fig:fig4): Barplot of Total DEGs

### Male and Female samples processed separately with TEs and outlier libraries removed
- #### Figure \@ref(fig:fig5): MDS plot of female samples only

- #### Figure \@ref(fig:fig6): MDS plot of male samples only

- #### Figure \@ref(fig:fig7): Barplot of Total DEGs

@@

@@

##Clean Trait Data and calculate CPMs
countdata=read.csv("/Users/johncsantiago/Documents/GitHub/WhartonLab/TKT_RNAseq/CountTables/TKT.counttable.csv",row.names=1)
groups=read.csv("/Users/johncsantiago/Documents/GitHub/WhartonLab/TKT_RNAseq/CountTables/TKT.metadata.csv",row.names=1)

write.csv(all.groups, "/Users/johncsantiago/Documents/GitHub/WhartonLab/TKT_RNAseq/CountTables/TKT.metadata.csv")

@@

@@

```{r fig1, echo = F, warning=F, fig.height=10, fig.width=10, fig.cap= 'MDS plot, all samples.'}

@@

```{r fig2, echo = F, fig.cap='Barplot of Total DEGs, all samples.'}

@@

```{r, include = F}
###EdgeR with TEs and outlier libraries removed

@@

```{r fig3, echo = F, warning=F, fig.height=10, fig.width=10, fig.cap= 'MDS plot, all samples, TEs and outlier libraries removed.'}

@@

```{r fig4, echo = F, fig.cap='Barplot of Total DEGs, all samples, TEs and outlier libraries removed.'}

@@

@@

```{r fig5, echo = F, warning=F, fig.height=10, fig.width=10, fig.cap= 'MDS plot, females only, TEs and outliers removed.'}

@@

```{r fig6, echo = F, warning=F, fig.height=10, fig.width=10, fig.cap= 'MDS plot, males, TEs and outlier libraries removed.'}

@@

```{r fig7, echo = F, fig.cap='Barplot of Total DEGs, males and females processed separately, TEs and outlier libraries removed.'}

@@

```{r echo = F, fig.cap='Barplot of Total DEGs excluding TEs.'}

@@

fig <- fig %>% add_trace(y = ~Female.trim, name = 'Trim Female',
                         marker = list(color = c("pink"),
                                       line = list(color = "black", width = 1.5)))

fig <- fig %>% add_trace(y = ~Female.Only.trim, name = 'Trim Only Female',
                         marker = list(color = c("red"),
                                       line = list(color = "black", width = 1.5)))

fig <- fig %>% add_trace(y = ~Male.trim, name = 'Trim Males',
                         marker = list(color = c("skyblue"),
                                       line = list(color = "black", width = 1.5)))

fig <- fig %>% add_trace(y = ~Male.Only.trim, name = 'Trim Only Males',
                         marker = list(color = c("blue"),
                                       line = list(color = "black", width = 1.5)))

>>>>

### Testing potential outlier samples

- #### Figure \@ref(fig:fig3): PCA of GR.F CxDf DEGs

- #### Figure \@ref(fig:fig4): PCA of GR.F CxOE DEGs

- #### Figure \@ref(fig:fig5): PCA of GR.FC x WT.FC DEGs

- #### Figure \@ref(fig:fig6): PCA of WT.F CxDf DEGs

- #### Figure \@ref(fig:fig7): PCA of WT.F CxOE DEGs

- #### Figure \@ref(fig:fig8): PCA of TktDfGR.F x TktDfWT.F DEGs

### Male and female samples processed together with TEs and outlier libraries removed
- #### Figure \@ref(fig:fig9): MDS plot, all samples

- #### Figure \@ref(fig:fig10): Barplot of Total DEGs

### Male and Female samples processed separately with TEs and outlier libraries removed
- #### Figure \@ref(fig:fig11): MDS plot of female samples only

- #### Figure \@ref(fig:fig12): MDS plot of male samples only

- #### Figure \@ref(fig:fig13): Barplot of Total DEGs

- #### Figure \@ref(fig:fig14): Barplot of Total DEGs for each comparison under the different trimming conditions

@@

### Conclusion: GR.F3, WT.F3 and TktDfWT.F3 are outlier libraries that should be removed before further analysis but there is no need to process male and female samples separately

<br /><br /><br /><br /><br /><br />
\newpage

@@

##Clean Trait Data and calculate CPMs
countdata=read.csv("/Users/johncsantiago/Documents/GitHub/WhartonLab/TKT_RNAseq/CountTables/TKT.counttable.csv",row.names=1)
groups=read.csv("/Users/johncsantiago/Documents/GitHub/WhartonLab/TKT_RNAseq/CountTables/All.TKT.metadata.csv",row.names=1)

@@

GroupsToDEGs = matrix("", nrow = 28, ncol = 3)

GroupsToDEGs[,1] = c("GR.F", "TktDfGR.F",
                     "GR.F", "TktOEGR.F",
                     "WT.F", "TktDfWT.F",
                     "WT.F", "TktOEWT.F",
                     "GR.F", "WT.F",
                     "TktDfGR.F", "TktDfWT.F",
                     "TktOEGR.F", "TktOEWT.F",
                     "GR.M", "TktDfGR.M",
                     "GR.M", "TktOEGR.M",
                     "WT.M", "TktDfWT.M",
                     "WT.M", "TktOEWT.M",
                     "GR.M", "WT.M",
                     "TktDfGR.M", "TktDfWT.M",
                     "TktOEGR.M", "TktOEWT.M")

GroupsToDEGs[,2] = c("TktDfGR.F", "GR.F",
                     "TktOEGR.F", "GR.F", 
                     "TktDfWT.F", "WT.F", 
                     "TktOEWT.F", "WT.F", 
                     "WT.F", "GR.F", 
                     "TktDfWT.F", "TktDfGR.F", 
                     "TktOEWT.F", "TktOEGR.F", 
                     "TktDfGR.M", "GR.M", 
                     "TktOEGR.M", "GR.M", 
                     "TktDfWT.M", "WT.M", 
                     "TktOEWT.M", "WT.M", 
                     "WT.M", "GR.M", 
                     "TktDfWT.M", "TktDfGR.M", 
                     "TktOEWT.M", "TktOEGR.M")

GroupsToDEGs[,3] = c(1,1,2,2,3,3,4,4,5,5,6,6,7,7,
                     8,8,9,9,10,10,11,11,12,12,13,13,14,14)

@@

## <span style="text-decoration:underline">Do the samples group by condition?</span> 
```{r fig1, echo = F, warning=F, fig.height=10, fig.width=10, fig.cap= 'MDS plot using all samples. Of these, samples GR.F3, WT.F3 and TktDfWT.F3 stand out as potential outliers.'}

@@

<br /><br /><br /><br /><br /><br />
\newpage

```{r fig2, echo = F, fig.cap='Barplot of Total DEGs observed with edgeR when using all samples.'}

@@

<br /><br /><br /><br /><br /><br />
\newpage

```{r, echo =F, include=F}
##Supervised PCAs

##Samples to plot
##genotype = "GR"                       ##options = "GR", "WT", or "all"
##sex      = "F"                         ##options = "M", "F", or "all"
##tkt      = c("Control", "DF")                    ##options = "Control", "DF", "OE", or "all"

                                       ##must choose genotype of first and second condition
##DEGs.genotypes = c("GR", "GR")         ##length = 2; options = "GR", "WT", or "all"

                                       ##a single sex, won't compare between sexs
##DEGs.sex       = "F"                   ##length = 1; options = "M", "F"

                                       ##choose tkt expression of first and second condition
##DEGs.tkt       = c("Control", "DF")    ##length = 1; "Control", "DF", "OE"

                                  ##must choose genotype of first and second condition
genotypes = c("GR", "GR")         ##length = 2; options = "GR", "WT", or "all"

                                  ##a single sex, won't compare between sexs
sex       = "F"                   ##length = 1; options = "M", "F"

                                  ##choose tkt expression of first and second condition
tkt       = c("Control", "DF")    ##length = 1; "Control", "DF", "OE"

##SupervisedPCA = function(genotype, sex, tkt, DEGs.genotypes, DEGs.sex, DEGs.tkt){
SupervisedPCA = function(genotypes, sex, tkt){
Group1=unique(groups[(groups$Genotype == genotypes[1] & ##DEGs.genotypes[1] &
                      groups$Sex == sex & ##DEGs.sex &
                      groups$TKT == tkt[1]), ##DEGs.tkt[1]), 
                      "Group"])

Group2=unique(groups[(groups$Genotype == genotypes[2] & ##DEGs.genotypes[2] &
                      groups$Sex == sex & ##DEGs.sex &
                      groups$TKT == tkt[2]), ##DEGs.tkt[2]), 
                      "Group"])

DEGs = GroupsToDEGs[(GroupsToDEGs[,1] == Group1 &
                     GroupsToDEGs[,2] == Group2),
                    3]

DEGs = Comparisons[[as.numeric(DEGs)]]

    use.groups =  groups[groups$Group == Group1 |
                         groups$Group == Group2,]

    ##if(genotype == "GR" | genotype == "WT"){
      ##use.groups = use.groups[use.groups$Genotype == genotype,]
    ##}

    ##if(sex == "M" | sex == "F"){
      ##use.groups = use.groups[use.groups$Sex == sex,]
    ##}

    ##if(length(tkt) == 1 & length(intersect(tkt, "all")) == 0){
      ##use.groups = use.groups[use.groups$TKT == tkt,]
    ##}

    ##if(length(tkt) == 2){
      ##use.groups = use.groups[use.groups$TKT == tkt[1]|
                                ##use.groups$TKT == tkt[2],]
    ##}

    use.data   = cpmdata[DEGs,row.names(use.groups)]
    pca <- prcomp(t(use.data), scale.=TRUE) 
    gr <- as.data.frame(use.groups)
    PC1=scale(pca$x[,1])
    PC2=scale(pca$x[,2])
    eigs <- pca$sdev^2
    ve=signif(((eigs / sum(eigs))*100)[1:3],4)

    pca.data=data.frame(Sample=row.names(gr),
                        PC1=scale(pca$x[,1]),
                        PC2=scale(pca$x[,2]),
                        sdev=pca$sdev,
                        Genotype=gr$Genotype, 
                        Sex=gr$Sex, 
                        TKT=gr$TKT,
                        Percent1=ve[1],
                        Percent2=ve[2],
                        Group=gr$Group)

  fig  = plot_ly(data = pca.data, 
                 x = ~PC1, 
                 y = ~PC2,
                 name = ~Group,
                 type = 'scatter',
                 mode = 'markers',
                 showlegend = T,
                 color = ~Group,
                 symbol = ~TKT,
                 ##symbols = as.character(seq_along(unique(point.data$Set))),
                 colors = c("firebrick","gold","deepskyblue"),
                 marker = list(size = 25,
                               line = list(color = "black", width = 1.5)),
                 hoverinfo = "text",
                 hovertext = paste("Sample:", pca.data$Sample)) %>%
    layout(xaxis = list(title = paste0("PC1", "(", ve[1], "%)")),
           yaxis = list(title = paste0("PC2", "(", ve[2], "%)")))
  return(fig)
  }

## <span style="text-decoration:underline">Is the expression profile for GR.F3 an outlier in genes found DE GR.FC and GR.FDf?</span> 
```{r fig3, echo = F, warning=F, fig.height=8, fig.width=8, echo = F, fig.cap='PCA of GR.F CxDf DEGs. Testing if GR.F3 is an outlier library.'}
##Supervised PCAs
##Samples to plot
                                  ##must choose genotype of first and second condition
genotypes = c("GR", "GR")         ##length = 2; options = "GR", "WT", or "all"

                                  ##a single sex, won't compare between sexs
sex       = "F"                   ##length = 1; options = "M", "F"

                                  ##choose tkt expression of first and second condition
tkt       = c("Control", "DF")    ##length = 1; "Control", "DF", "OE"

SupervisedPCA(genotypes, sex, tkt)

<br /><br /><br /><br /><br /><br />
\newpage

## <span style="text-decoration:underline">Is the expression profile for GR.F3 an outlier in genes found DE GR.FC and GR.FOE?</span> 
```{r fig4, echo = F, warning=F, fig.height=8, fig.width=8, fig.cap='PCA of GR.F CxOE DEGs. Testing if GR.F3 is an outlier library.'}
##Supervised PCAs
##Samples to plot
                                  ##must choose genotype of first and second condition
genotypes = c("GR", "GR")         ##length = 2; options = "GR", "WT", or "all"

                                  ##a single sex, won't compare between sexs
sex       = "F"                   ##length = 1; options = "M", "F"

                                  ##choose tkt expression of first and second condition
tkt       = c("Control", "OE")    ##length = 1; "Control", "DF", "OE"

SupervisedPCA(genotypes, sex, tkt)

<br /><br /><br /><br /><br /><br />
\newpage

## <span style="text-decoration:underline">Are the expression profiles for GR.F3 and WT.F3 outliers in genes found DE GR.FC and WT.FC?</span> 
```{r fig5, echo = F, warning=F, fig.height=8, fig.width=8, fig.cap='PCA of GR.FC x WT.FC DEGs. Testing if GR.F3 and WT.F3 are outlier libraries.'}
##Supervised PCAs
##Samples to plot
                                  ##must choose genotype of first and second condition
genotypes = c("GR", "WT")         ##length = 2; options = "GR", "WT", or "all"

                                  ##a single sex, won't compare between sexs
sex       = "F"                   ##length = 1; options = "M", "F"

                                  ##choose tkt expression of first and second condition
tkt       = c("Control", "Control")    ##length = 1; "Control", "DF", "OE"

SupervisedPCA(genotypes, sex, tkt)

<br /><br /><br /><br /><br /><br />
\newpage

## <span style="text-decoration:underline">Are the expression profiles for WT.F3 and TktDfWT.F3 outliers in genes found WT.FC and TktDfWT.F?</span> 
```{r fig6, echo = F, warning=F, fig.height=8, fig.width=8, fig.cap='PCA of WT.F CxDf DEGs. Testing if WT.F3 and TktDfWT.F3 are outlier libraries.'}
##Supervised PCAs
##Samples to plot
                                  ##must choose genotype of first and second condition
genotypes = c("WT", "WT")         ##length = 2; options = "GR", "WT", or "all"

                                  ##a single sex, won't compare between sexs
sex       = "F"                   ##length = 1; options = "M", "F"

                                  ##choose tkt expression of first and second condition
tkt       = c("Control", "DF")    ##length = 1; "Control", "DF", "OE"

SupervisedPCA(genotypes, sex, tkt)

<br /><br /><br /><br /><br /><br />
\newpage

## <span style="text-decoration:underline">Is the expression profile for WT.F3 an outlier in genes found DE WT.FC and WT.FOE?</span> 
```{r fig7, echo = F, warning=F, fig.height=8, fig.width=8, fig.cap='PCA of WT.F CxOE DEGs. Testing if WT.F3 is an outlier library.'}
##Supervised PCAs
##Samples to plot
                                  ##must choose genotype of first and second condition
genotypes = c("WT", "WT")         ##length = 2; options = "GR", "WT", or "all"

                                  ##a single sex, won't compare between sexs
sex       = "F"                   ##length = 1; options = "M", "F"

                                  ##choose tkt expression of first and second condition
tkt       = c("Control", "OE")    ##length = 1; "Control", "DF", "OE"

SupervisedPCA(genotypes, sex, tkt)

<br /><br /><br /><br /><br /><br />
\newpage

## <span style="text-decoration:underline">Is the expression profile for TktDfWT.F3 an outlier in genes found DE TktDfGR.F and TktDfWT.F?</span> 
```{r fig8, echo = F, warning=F, fig.height=8, fig.width=8, fig.cap='PCA of TktDfGR.F x TktDfWT.F DEGs. Testing if TktDfWT.F3 is an outlier library.'}
##Supervised PCAs
##Samples to plot
                                  ##must choose genotype of first and second condition
genotypes = c("GR", "WT")         ##length = 2; options = "GR", "WT", or "all"

                                  ##a single sex, won't compare between sexs
sex       = "F"                   ##length = 1; options = "M", "F"

                                  ##choose tkt expression of first and second condition
tkt       = c("DF", "DF")    ##length = 1; "Control", "DF", "OE"

SupervisedPCA(genotypes, sex, tkt)

<br /><br /><br /><br /><br /><br />
\newpage

```{r, echo = F, include = F}
###EdgeR with TEs and outlier libraries removed

@@

## <span style="text-decoration:underline">Does processing the samples after removing the 3 outlier libraries and TEs improve sample normalization?</span> 
```{r fig9, echo = F, warning=F, fig.height=10, fig.width=10, fig.cap= 'MDS plot using all samples, with the TEs and outlier libraries removed.'}

@@

<br /><br /><br /><br /><br /><br />
\newpage

```{r fig10, echo = F, fig.cap='Barplot of Total DEGs found when using all samples with the TEs and outlier libraries removed.'}

@@

<br /><br /><br /><br /><br /><br />
\newpage

@@

## <span style="text-decoration:underline">Does sex specifically processing the samples after removing the 3 outlier libraries and TEs improve sample normalization?</span> 
```{r fig11, echo = F, warning=F, fig.height=10, fig.width=10, fig.cap= 'MDS plot when the female samples are processed alone. The TEs and outlier libraries are removed.'}

@@

```{r fig12, echo = F, warning=F, fig.height=10, fig.width=10, fig.cap= 'MDS plot when the male samples are process alone only. The TEs are removed.'}

@@

<br /><br /><br /><br /><br /><br />
\newpage

```{r fig13, echo = F, fig.cap='Barplot of Total DEGs when male and female samples are processed separately. The TEs and outlier libraries were removed.'}

@@

<br /><br /><br /><br /><br /><br />
\newpage

## <span style="text-decoration:underline">How do the different processing methods alter the number of total DEGs?</span> 
```{r fig14, echo = F, fig.cap='Barplot of Total DEGs for each comparison under the different trimming conditions. Different trimming conditions were: (1) all samples processed together with no TEs or outlier libraries removed; (2) trimmed data with TEs and outlier libraries removed, male and female samples processed together; (3) trimmed data with TEs and outlier libraries removed, male and female samples processed separately.'}

@@

fig <- fig %>% add_trace(y = ~Female.trim, name = 'Trim Female',
                         marker = list(color = c("red"),
                                       line = list(color = "black", width = 1.5)))

fig <- fig %>% add_trace(y = ~Female.Only.trim, name = 'Trim Only Female',
                         marker = list(color = c("pink"),
                                       line = list(color = "black", width = 1.5)))

fig <- fig %>% add_trace(y = ~Male, name = 'Males',
                         marker = list(color = c("blue"),
                                       line = list(color = "black", width = 1.5)))

fig <- fig %>% add_trace(y = ~Male.trim, name = 'Trim Males',
                         marker = list(color = c("cornflowerblue"),
                                       line = list(color = "black", width = 1.5)))

fig <- fig %>% add_trace(y = ~Male.Only.trim, name = 'Trim Only Males',
                         marker = list(color = c("skyblue"),
                                       line = list(color = "black", width = 1.5)))
